dyngen: a multi-modal simulator for spearheading new single-cell omics analyses

### 2020-06-13 manuscript.Rmd
update

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- Robrecht Cannoodt*
- Wouter Saelens*
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

\printbibliography

# Supplementary Figures

\newrefsection
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[H]
    \centering
    \includegraphics[width=.8\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
 \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}. In \textbf{C}, the individual mappings of the alignment between cells are shown. In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix. In E the complete count matrices were used to perform the alignment. In F, smoothed pseudocells were used.
  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
    }
    \label{fig:traj_align}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{result_files/explain_methods}
    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network. Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. 
    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the molecules at each time step are extracted. Each time step is mapped to a point in the backbone. 
    \textbf{F:} The molecule levels of multiple simulations are shown over time (left). From each simulation, multiple cells are sampled (from left to middle). Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
    }
    \label{fig:explain_methods}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
    \caption{
      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
    }
    \label{fig:example_backbones_onlymodules}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
    \caption{
        \textbf{Generating the feature network from a backbone consists of four main steps.}
    }
    \label{fig:gen_feature_network}
\end{figure}

@@

\begin{table}[ht]
    \caption{
      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
    } \label{tab:reaction_def}
    \centering
    \begin{tabular}{|lcc|}
        \hline
        Reaction & Effect & Propensity \\ \hline \hline
        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
    \end{tabular}
\end{table}

\begin{table}[ht]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_params}
    \centering
    \begin{tabular}{|lrl|}
        \hline
        Parameter & Symbol & Definition \\ \hline \hline
        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
    \caption{
        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
    }
    \label{fig:sample_cells}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
    \label{fig:example_backbones}
\end{figure}

@@

Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN <!--(Figure\ \ref{fig:cell-specific_grn})-->. The regulatory effect lies between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
<!--
\begin{figure}[htb!]
    \centering
    \includegraphics[width=.8\linewidth]{result_files/fig1_showcase_functionality/scgrn.pdf}
    \caption{
        \textbf{The cell-specific regulatory effects of all interactions, computed on cells part of a bifurcation trajectory.} Negative values correspond to inhibitory interactions, positive values to activating interactions, and zero values correspond to inactive interactions. 
    }
    \label{fig:cell-specific_grn}
\end{figure}
-->

@@

>>>>

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- Robrecht Cannoodt*
- Wouter Saelens*
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

\printbibliography
\newpage

@@

@@

@@

@@

@@

@@

@@

Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN. The regulatory effect lies between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.

@@

\printbibliography
\newpage

# Supplementary Figures
\setcounter{page}{1}
\newrefsection
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
 \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}. In \textbf{C}, the individual mappings of the alignment between cells are shown. In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix. In E the complete count matrices were used to perform the alignment. In F, smoothed pseudocells were used.
  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
    }
    \label{fig:traj_align}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{result_files/explain_methods}
    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network. Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. 
    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the molecules at each time step are extracted. Each time step is mapped to a point in the backbone. 
    \textbf{F:} The molecule levels of multiple simulations are shown over time (left). From each simulation, multiple cells are sampled (from left to middle). Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
    }
    \label{fig:explain_methods}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
    \caption{
      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
    }
    \label{fig:example_backbones_onlymodules}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
    \caption{
        \textbf{Generating the feature network from a backbone consists of four main steps.}
    }
    \label{fig:gen_feature_network}
\end{figure}

\begin{table}[H]
    \caption{
      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
    } \label{tab:reaction_def}
    \centering
    \begin{tabular}{|lcc|}
        \hline
        Reaction & Effect & Propensity \\ \hline \hline
        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
    \end{tabular}
\end{table}

\begin{table}[H]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_params}
    \centering
    \begin{tabular}{|lrl|}
        \hline
        Parameter & Symbol & Definition \\ \hline \hline
        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
    \caption{
        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
    }
    \label{fig:sample_cells}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
    \label{fig:example_backbones}
\end{figure}
